Scheduling Musical Events in Max/MSP with Scheme For Max

### 2022-10-07 papers/paper-1/paper-1.tex
updating structure

As externals are compiled plug-ins, they do not need to be distributed as source-code, and
are most typically made available as binaries for Windows and OSX. Extending Max through externals has 
been possible since very early versions of Max in the late 1980's,
thus thousands of 3rd party externals 
now exist \cite{MaxObjects2021}, both open and closed source, of which Scheme For Max is one.

@@

Scheme For Max is far from the first Lisp-based computer music tool, or even
the first real-time music tool in Lisp. There is a rich history of Lisp in music, 
ranging from Common Lisp Music, created by Bill Schottstaedt in the late 1980's
\cite{Wang17} and Heinrich Taube's Common Music in 1991 \cite{Taube91},
to Dannenberg's Nyquist in 1997 \cite{Dannenberg97},
and Impromptu (later rewritten as Extempore) \cite{Sorensen2010}. 
Nor is Scheme For Max unique in providing a Lisp-family
extension facility to a larger music platform --- the author unknowingly
programmed in Lisp in the early 1990's while using the Cakewalk sequencer,
which came with an embedded Lisp interpreter in the form of the 
Cakewalk Application Language (CAL)! There also exist Lisp-family client interfaces
to the open-source SuperCollider music platform, including Overtone (Clojure),
rsc3 (r6rs Scheme), and cl-collider (Common Lisp).

@@

In contrast to these, S4M embeds a light-weight Scheme interpreter directly in a Max
C external with no intermediate or external runtime necessary. This is made possible by the suitability
of Scheme for the creation of small, portable, and self-contained interpreters.
The s7 interpreter is implemented entirely in ANSI C, is statically linked during compilation of S4M, 
and operates entirely in whatever thread it is run in by the C host.
This means multiple instantiations of the interpreter are possible, and
each \textbf{s4m} object has direct C-level access to the Max C API 
through foreign function interface calls, with the results of said calls being exactly 
the same as if the functions were called in C, save execution time. 
The ramification of this is that S4M is unique among Max Lisp projects in being
able to operate within the native Max scheduling system, enabling highly accurate
timing and retaining deterministic control flow within a patcher. (i.e., There
is no hidden networking layer with communication to an outside process, which 
makes operations fundamentally asynchronous even if this is hidden from the user visually.)
This provides the user with opportunities to use Scheme for tasks beyond
what is possible with pre-existing options. For example, we are in effect 
able to say ``at a certain time, read some shared state and update an 
internal engine parameter accordingly". In the example below, a Scheme function
is scheduled that will, in 4 bars time, read a rehearsal mark from the table \textbf{mark-table},
advancing the global transport to that position. As the table
is a shared data object, other Max objects (or patchers) could write to this
rehearsal mark table without awareness of S4M. (Note: this is a safe operation
from Scheme code as table access is done through thread-safe Max SDK functions 
that handle synchronization and locking for us.)

@@

The s7 interpreter is implemented in a single C file (with header file), 
and is statically linked with the C code of the \textbf{s4m} object. Embedding is
as simple as linking to the s7 library and including a single file. In this 
regard, it is similar to Guile, and in fact, is linguistically similar to Guile 1.8.
However, while Guile is reciprocally licensed with the GPL, s7 is liberally licensed 
under the BSD license. As the creation and sale of commercially externals is part of the Max 
milieu, the author felt this was more likely to result in uptake within the Max community.

@@

Finally, s7 was created for computer music, and as a result is used in other computer music
environments, such as Common Music. As a result, there is a rich body of code, especially
from Common Music, that can be used or studied by users. There is also a small, but active, 
online community of music programmers using s7, communicating through the CM-dist mailing
list hosted at the Stanford computer music center, CCRMA. The author of s7, Bill
Schottstaedt, has been an enthusiastic supporter of the project and has been invaluable
with technical assistance.

@@

In addition to both of these, the \textbf{s4m} object implements a \textbf{read} message,
where sending \textbf{read filename} to the object will result in the full filename
being found on the Max search path and used in a call to the Scheme \textbf{load} function,
without resetting the s7 interpreter.

@@

For performers and composers exploring the intersections of live performance and algorithmic
composition, one is ideally able to use gestural inputs not just to affect
what is happening \textit{now} (as with a traditional musical instrument), but also what will 
happen \textit{in the future}. For example, a performer may have two physical dials,
and may wish to schedule an event in the future that will use parameters derived
from these dials, but may wish one parameter to be used as the dial is now
(at the time of event dispatch) 
and the other to be used as the performer will have the dial at the scheduled time.
Further, the absolute time of the event may not even be known at the time of 
scheduling, as could be the case if the event time was specified in musical terms 
(i.e., on the down-beat of the next 8 bar section) and the tempo might be changed in real-time.

@@

A larger problem is that of how to capture gestural values for use in the delayed
events. Max's visual patching language is fundamentally modelled similarly to a modular
synthesizer --- there is one instance in memory of each visual object, and messages can only
pass through them one at a time. Creating visual programs where execution of 
delayed events will trigger cascades of messages through objects that are also 
being used in the interim becomes notoriously complex and there exists no straightforward
facility to express \textit{``make a new container for this variable as it is now so that
we can fetch it later"}. A naive solution would use a visual patcher object to store 
the variable's state at trigger time for use later, however this only works
if there is only one scheduled event --- adding more scheduled events requires adding
an additional storage objects per concurrently scheduled event, and as each object in the visual patcher
represents an object instantiation rather than a class, this rapidly becomes
baroque. This ``programming-with-instances" paradigm
is very convenient when we want to ensure there is only ever one instance of a given function, such
as in emulation of a hardware synthesizer, where each oscillator represents one ``always-on"
physical device that is playing regardless of whether one can hear it. 
However, it is difficult in a polyphonic time context
such as the creation and eventual playback of some arbitrary number of delayed events.

@@

It should also be mentioned that there exists within Max a facility for creating
new instances of Max sub-patches on demand, designed with the express purpose of
overcoming this barrier. One can use the \textbf{poly} object to create polyphonic
patches. However, this involves complex patching, so we will not examine this further here,
save to comment that this again becomes very cumbersome when dealing
with a potentially large and unknown number of events, each with arbitrary numbers
of parameters.

In the C API however, Max does provide a highly accurate facility for delaying
the execution of a C callback function, through the Max \textbf{clock} API. Scheme
for Max builds on this to provide an accurate, flexible, and high-level means 
for scheduling future events, without the problems described.

@@

We can see that the visual patching language of Max is not well suited to implementing
our use case of a performer wanting to schedule large numbers of events with arbitrary numbers of
parameters and gestural inputs meant variously for present and future use. 
However, Max does provide an alternative programming solution through its
\textbf{js} object, which embeds a JavaScript (ECMA5) interpreter in the Max environment.
The user is able to load JavaScript code from a file given as an argument to the object,
and this code can interact with the Max environment and with object inlets and outlets,
similar to how one can do so in C code in a Max external.
This does allow one to create functions and variables, and does have a facility
for delayed execution through the JavaScript \textbf{Task} object. 
Thus technically it could be used to solve our use case.
In many respects the \textbf{js} object is functionally similar to the \textbf{s4m} object,
and in fact, the desire to work with Max this way, combined with dissatisfaction with the
limitations of the Max JavaScript implementation were the reasons for the development of S4M.

The \textbf{js} object in Max is not an ideal solution to our problem for several reasons. 
The most serious is that it is limited to running only in the Max low-priority GUI thread \cite{Cycling74}.
As this thread is also used for Max file i/o and graphic redrawing, the result is that latency
is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
The human ear is very sensitive to time, with delays of tens of milliseconds enough to sound like errors in playback
of highly rhythmic music. This means that it takes little other activity in the
GUI thread to delay our scheduled events enough to be audibly incorrect. 
For some purposes, this is acceptable, but for the creation of accurate 
musical sequencers or algorithmic music engines, the \textbf{js} object is unreliable.

@@

\begin{itemize}
\item A data structure to hold state used by the object
\item A class building function used to create the class in C
\item An instance constructor function called when objects are added to a patch
\item Any methods that will be bound to messages as event handlers
\end{itemize}

@@

Let us imagine instead that the \textbf{bang} message should delay the activity
(posting to the console) by 1000 ms. We can use the Max \textbf{clock} facility for high-accuracy
delay as it accepts floats for the ms time value. (Note that the actual delay time
will be offset to the nearest signal vector boundary, determined by the
users ``Signal Vector Size" audio setting, but subsequent clock calls 
adjust for this, maintaining long-term temporal accuracy at sample rate resolution.)

We will create a \textbf{clock} object, which takes as arguments a void pointer and a function
reference, with the void pointer normally used to hold a reference to our instantiated
object (x). The standard method for doing this in Max (taken from the SDK documentation)
is to add a \textbf{clock} object to our classes data structure, and to add the act
of starting the clock to the \textbf{bang} handler.

@@

From the above we can see that, while accurate, the clock functionality is limited -
the callback must be a single-arity function expecting a pointer, and this
is normally to the object.

@@

Scheme For Max builds on the clock facility provided by the Max API to allow scheduling
Scheme functions. The delayed functions are limited to zero-arity signatures, but as
creation of lambda functions in Scheme is trivial, this is of no practical significance to 
user. To the user of S4M, scheduling the execution of a function is simple:

@@

Cancelling a delay function consists of merely replacing the callback registered
in the hash-table with the value \textbf{\#false}, letting the clock fire harmlessly.
The s7 Scheme code for this is shown below. It uses an s7's \textbf{gensym} function to create the
symbolic handle that is guaranteed to be unique to this instance of the interpreter.

@@

\begin{itemize}
\item A \textbf{t\_s4m\_clock\_callback} data structure is dynamically allocated and used to store a
  reference to the \textbf{s4m} object and to the symbolic handle passed from Scheme.
\item A Max \textbf{clock} is created, passing in a void pointer to clock callback structure
  and binding the clock to a generic clock callback function in C. The clock's timer is started.
\item The clock callback struct is also stored in a C hash-table (Max's hashtab implementation),
  keyed by the handle, so that cancellation or object deletion can use this to find all clocks.
\item When the generic clock callback function runs (time has elapsed), it uses the pointer argument to get 
  the delay handle and the instantiated \textbf{s4m} object, through which it 
  can also get the s7 interpreter pointer.
\item The s7 interpreter and handle are then used to call the Scheme function
  \textbf{s4m-execute-callback,} which will run our delayed Scheme function as previously explained.
\item The generic C callback then removes the clock reference from the C hash-table,
  deletes the clock, and frees the memory allocated for the callback struct.
\end{itemize}

@@

In addition to output latency, there is the factor of internal latency.
Max, like many computer music programming environments, generates audio 
in vectors of samples, with
each vector generated from one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass \cite{Puckette2002}.
The size of this signal vector thus has a large effect on performance, with larger
vectors requiring significantly less CPU use. As audio rendering of note onsets originating
from control messages (the normal scenario) must thus be aligned with 
signal vector boundaries, this produces delays of note onsets by up to
the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
However, the more recent scheduling facilities in Max, including the 
clock facilities used by S4M, are implemented to compensate for this jitter,
maintaining long-term temporal accuracy despite note onset delays. The mechanism
by which this is done is unfortunately not publicly documented or available
in open-source code, but one can assume that it is some variant of the common pattern 
whereby scheduled events are aware of when they are supposed to be running
(``logical-time"), and if they in turn schedule subsequent events, the time for these subsequent
events will be shifted to compensate for the jitter, based 
on the logical time of the scheduling event, rather than its
actual (post-jitter) real time \cite{Anderson1986}.
The author also conducted personal tests of 
this compensation, comparing both high and low signal vector sizes to produce
variable delays in note onsets, and it was determined
that so long as the overall output latency is again large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that jitter does not accumulate over the piece --- 
note onsets stay accurate to within the delay of one signal vector of samples,
regardless of the length of the piece of music.

@@

Future plans for the project include a version for the open-source
Pure Data platform, similar to Max in many ways and created also by Miller Puckette,
the original author of Max. \textbf{Scheme For Pure Data} is now in beta release, available
as source code, with binary releases planned for August 2021. In addition
to this, experimental work on running Scheme for control-rate DSP is planned.
This would entail Scheme executing in the timed audio rendering thread, but
at a subdivision of the audio sample rate such as every 32 or 64 samples, 
thus generating streams similar to the way digital audio is rendered, but
at a lower resolution suitable for use as signal modulators. (The ability
to run DSP code at a lower sample rate for modulator signals 
is called the ``control rate" in CSound and other similar audio programming platforms.)

The author also plans to create portability layers in Scheme to allow one to run 
the same Scheme code on S4M, S4Pd, or Common Music (which also uses s7),
and potentially in C++ projects through the JUCE audio development framework.
There have also been successful ports of s7, Pure Data, and Csound to Web Assembly, 
thus it seems likely that work from the project will be useable for web audio as well.
It is hoped that through these additional run-times, the project can become 
a valuable addition to the larger computer music community, enabling
use in contexts where the commercial Max platform is not practical
or financially accessible.

@@

https://github.com//scheme-for-max
%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
Scheme For Max would not exist without the work of many people. 
Bill Schottstaedt created Common Lisp Music and the s7 Scheme implementation,
and has been invaluable as a source of advice on s7. Heinrich Taube
created Common Music and has been an inspiration through his seminal book
on algorithmic composition with Lisp, "Notes from the Metalevel".
Luigi Castelli and Alex Harker helped with programming questions
and build assistance. 
S4M of course depends on Max and thus owes a huge debt to 
Miller Puckette and David Zicarelli.
George Tzanetakis has assisted with this paper.
And finally, Andrew Schloss has
been a source of encouragement and support through the years of 
exploration that led to the Scheme For Max project.

>>>>

As externals are compiled plug-ins, they do not need to be distributed as source-code, and
are most typically made available as binaries for Windows and macOS. Extending Max through externals has 
been possible since very early versions of Max in the late 1980's,
thus thousands of 3rd party externals 
now exist \cite{MaxObjects2021}, both open and closed source, of which Scheme For Max is one.

@@

Scheme For Max is far from the first Lisp-based computer music tool, or even
the first real-time music tool in Lisp. There is a rich history of Lisp in music, 
including (but not limited to) Common Lisp Music, created by Bill Schottstaedt in the late 1980's
\cite{Wang17}, Heinrich Taube's Common Music from 1991 \cite{Taube91},
Roger Dannenberg's Nyquist from 1997 \cite{Dannenberg97},
and Andrew Sorensen's Impromptu from 2005 (later rewritten as Extempore) \cite{Sorensen2010}. 
Nor is Scheme For Max unique in providing a Lisp-family
extension facility to a larger music platform --- the author unknowingly
programmed in Lisp in the early 1990's while using the Cakewalk sequencer,
which came with an embedded Lisp interpreter in the form of the 
Cakewalk Application Language! There also exist Lisp-family client interfaces
to the open-source SuperCollider music platform, including Overtone (Clojure),
rsc3 (r6rs Scheme), and cl-collider (Common Lisp).

@@

In contrast to these, S4M embeds a light-weight Scheme interpreter directly in a Max
C external with no intermediate or external runtime necessary. This is made possible by the suitability
of Scheme for the creation of small, portable, and self-contained interpreters.
The s7 interpreter is implemented entirely in ANSI C, is statically linked during compilation of S4M, 
and operates entirely in whatever thread it is run in by the C host.
This means multiple instantiations of the interpreter are possible, and
each \textbf{s4m} object has direct C-level access to the Max C API 
through foreign function interface calls, with the results of said calls being exactly 
the same as if the functions were called in C, save execution time. 
The ramification of this is that S4M is unique among Max Lisp projects in being
able to operate within the native Max scheduling system, enabling highly accurate
timing and retaining deterministic control flow within a patcher. (i.e., There
is no hidden networking layer with communication to an outside process, which 
makes operations fundamentally asynchronous even if this is hidden from the user visually,
as is the case with Node for Max and MOZ'Lib.)
This provides the user with opportunities to use Scheme for tasks beyond
what is possible with previous options.

For example, we are in effect 
able to say ``at a certain time, read some shared state and update an 
internal engine parameter accordingly". In the example below, a Scheme function
is scheduled that will, in 4 bars time, read a rehearsal mark from the table \textbf{mark-table},
advancing the global transport to that position. As the table
is a shared data object, other Max objects (or patchers) could write to this
rehearsal mark table without awareness of S4M. (Note: this is a safe operation
from Scheme code as table access is done through thread-safe Max SDK functions 
that handle synchronization and locking for us.)

@@

The s7 interpreter is implemented in a single C file (with header file), 
and is statically linked with the C code of the \textbf{s4m} object. Embedding is
as simple as linking to the s7 library and including a single file. In this 
regard (ease of embedding), it is similar to Guile, and in fact, is linguistically similar to Guile 1.8.
However, while Guile is licensed with the (somewhat reciprocal) LGPL license, s7 is liberally licensed 
under the BSD license. As the creation and sale of commercially externals is part of the Max 
milieu, the author felt this was more likely to result in uptake within the Max community.

@@

Finally, s7 was created for computer music, and as a result is used in other computer music
environments, such as Common Music. As a result, there is a rich body of code, especially
from Common Music, that can be used or studied by users. There is also a small, but active, 
online community of music programmers using s7, communicating through the CM-dist mailing
list hosted at the Stanford computer music center, CCRMA. The author of s7, Bill
Schottstaedt, has been working with Lisp and music since the 1980's, and has been an 
enthusiastic supporter of the project, providing invaluable suggestions and 
technical assistance.

@@

In addition to both of these, the \textbf{s4m} object implements a \textbf{read} message,
where sending \textbf{read \{filename\} } to the object will result in the full filename
being found on the Max search path and used in a call to the Scheme \textbf{load} function,
without resetting the s7 interpreter.

@@

For performers and composers exploring the intersections of live performance and algorithmic
composition, one is ideally able to use gestural inputs not just to affect
what is happening \textit{now} (as with a traditional musical instrument), but also what will 
happen \textit{in the future}. For example, a performer may have two physical dials,
and may wish to schedule an event in the future that will use parameters derived
from these dials, but may wish one parameter to be used as the dial is now
(at the time of event dispatch) 
and the other to be used as the performer will have the dial at the scheduled time.
Further, the absolute time of the event may not even be known at the time of 
scheduling, as could be the case if the event time was specified in musical terms 
(i.e., on the down-beat of the next 8 bar section) and the tempo might be changed 
prior to the scheduled (musical) time.

@@

A larger problem is that of how to capture gestural values for use in the delayed
events. Max's visual patching language is fundamentally modelled similarly to a modular
synthesizer --- there is one instance in memory of each visual object, and messages can only
pass through them one at a time. Creating visual programs where execution of 
delayed events will trigger cascades of messages through objects that are also 
being used in the interim becomes onerously complex and there exists no straightforward
facility to express \textit{``make a new container for this variable as it is now so that
we can fetch it later"}. A naive solution would be to use a visual patcher object to store 
the variable's state at trigger time for use later, however this only works
if there is only one scheduled event --- adding more scheduled events requires adding
an additional storage objects per concurrently scheduled event, and as each object in the visual patcher
represents an object instantiation rather than a class, this rapidly becomes
baroque. This ``programming-with-instances" paradigm
is very convenient when we want to ensure there is only ever one instance of a given function, such
as in emulation of a hardware synthesizer, where each oscillator represents one ``always-on"
physical device that is playing regardless of whether one can hear it. 
However, it is difficult in a polyphonic time context
such as the creation and eventual playback of some arbitrary number of delayed events.

@@

It should also be mentioned that there exists within Max a facility for creating
new instances of Max sub-patches on demand, designed with the express purpose of
overcoming this barrier. One can use the \textbf{poly} object to create polyphonic
patches. However, this involves complex patching, so we will not examine this further here,
save to comment that this again becomes cumbersome when dealing
with a potentially large and unknown number of events, each with arbitrary numbers
of parameters.

In the C API however, Max does provide a highly accurate facility for delaying
the execution of a C callback function, through the Max \textbf{clock} API. Scheme
for Max builds on this to provide an accurate, flexible, and high-level means 
to schedule future events, without the problems described.

@@

We can see that the visual patching language of Max is not well suited to implementing
our use case of a performer wanting to schedule large numbers of events with arbitrary numbers of
parameters and gestural inputs meant variously for present and future use. 
However, Max does provide an alternative programming solution through its
\textbf{js} object, which embeds a JavaScript (ECMA5) interpreter in the Max environment.
The user is able to load JavaScript code from a file (provided as an argument to the object),
and this code can interact with the Max environment and with object inlets and outlets,
similar to how one can do so in C code in a Max external.
This does allow one to create functions and variables, and does have a facility
for delayed execution through the JavaScript \textbf{Task} object. 
Thus technically it could be used to solve our use case.
In many respects the \textbf{js} object is functionally similar to the \textbf{s4m} object,
and in fact, the desire to work with Max this way, but without the limitations of the Max
JavaScript implementation, was the reason for the development of S4M.

The \textbf{js} object in Max is not an ideal solution to our problem for several reasons. 
The most serious is that it is limited to running only in the Max low-priority GUI thread \cite{Cycling74}.
As this thread is also used for Max file i/o and graphic redrawing, the result is that latency
is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
The human ear is very sensitive to time, with delays of tens of milliseconds being enough to sound like errors in playback
of highly rhythmic music. This means that it takes little other activity in the
GUI thread to delay our scheduled events enough to be audibly incorrect. 
For some purposes, this is acceptable, but for the creation of accurate 
musical sequencers or algorithmic music engines, the \textbf{js} object is unreliable.

@@

\begin{itemize}
\item A data structure to hold state used by the object
\item A class-building function used to create the class in C
\item An instance constructor function called when objects are added to a patch
\item Any methods that will be bound to messages as event handlers
\end{itemize}

@@

Let us imagine instead that the \textbf{bang} message should delay the activity
of posting to the console by 1000 ms. We can use the Max \textbf{clock} facility for high-accuracy
delay, as it accepts floats for the ms time value. (Note that the actual delay time
will be offset to the nearest signal vector boundary, determined by the
users ``Signal Vector Size" audio setting, but subsequent clock calls 
adjust for this, maintaining long-term temporal accuracy. If one needs
single-sample accuracy, the solution is to simply set the signal vector to a size of 1.)

We will create a \textbf{clock} object, which takes as arguments a void pointer and a function
reference, with the void pointer normally used to hold a reference to our instantiated
object (x). The standard method for doing this in Max (taken from the SDK documentation)
is to add a \textbf{clock} object to our class's data structure, and to add the act
of starting the clock to the \textbf{bang} handler.

@@

From the above we can see that, while accurate, the clock functionality is limited ---
the callback must be a single-arity function expecting a pointer, and this
is normally a pointer to the object.

@@

Scheme For Max builds on the clock facility provided by the Max API to allow scheduling
Scheme functions. The delayed functions are limited to zero-arity signatures, but as
creation of lambda functions in Scheme is trivial, this is of little practical significance to 
user. To the user of S4M, scheduling the execution of a function is simple:

@@

Cancelling a delay function consists of merely replacing the callback registered
in the hash-table with the value \textbf{\#false}, letting the clock fire harmlessly.
The s7 Scheme code for this is shown below. It uses an s7's \textbf{gensym} function to create a
symbolic handle that is guaranteed to be unique to this instance of the interpreter.

@@

\begin{itemize}
\item A \textbf{t\_s4m\_clock\_callback} data structure is dynamically allocated and used to store a
  reference to the \textbf{s4m} object and to the symbolic handle passed from Scheme.
\item A Max \textbf{clock} is created, passing in a void pointer to a clock callback structure
  and binding the clock to a generic clock callback function in C. The clock's timer is started.
\item The clock callback struct is also stored in a C hash-table (using Max's \textbf{hashtab} implementation),
  keyed by the handle, so that cancellation or object deletion can use this to find all clocks.
\item When the generic clock callback function runs (time has elapsed), it uses the pointer argument to get 
  the delay handle and the instantiated \textbf{s4m} object, through which it 
  can also get the s7 interpreter pointer.
\item The s7 interpreter and handle are then used to call the Scheme function
  \textbf{s4m-execute-callback,} which will run our delayed Scheme function as previously explained.
\item The generic C callback then removes the clock reference from the C hash-table,
  deletes the clock, and frees the memory allocated for the callback struct.
\end{itemize}

@@

In addition to output latency, there is the factor of internal latency.
Max, like many computer music programming environments, generates audio 
in vectors of samples, with
each vector generated from one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass \cite{Puckette2002}.
The size of this signal vector thus has a large effect on performance, with larger
vectors requiring significantly less CPU use. As audio rendering of note onsets originating
from control messages (the normal scenario) must thus be aligned with 
signal vector boundaries, this produces delays of note onsets by up to
the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
However, the more recent scheduling facilities in Max, including the 
clock facilities used by S4M, are implemented to compensate for this jitter,
maintaining long-term temporal accuracy despite note onset delays. The mechanism
by which this is done is unfortunately not publicly documented or available
in open-source code, but one can assume that it is some variant of the common pattern 
whereby scheduled events are aware of when they are supposed to be running
(``logical-time"), and if they in turn schedule subsequent events, the time for these subsequent
events will be shifted to compensate for the jitter, based 
on the logical time of the scheduling event, rather than its
actual (post-jitter) actual clock time \cite{Anderson1986}.
The author also conducted personal tests of 
this compensation, comparing both high and low signal vector sizes to produce
variable delays in note onsets, and it was determined
that so long as the overall output latency is again large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that jitter does not accumulate over the piece (note onsets 
stay accurate to within the delay of one signal vector of samples,
regardless of the length of the piece of music).

@@

Future plans for the project include a version for the open-source
Pure Data platform, similar to Max in many ways and created also by Miller Puckette,
the original author of Max. \textbf{Scheme For Pure Data} is now in beta release, available
as source code, with binary releases planned for August 2021. In addition
to this, experimental work on running Scheme for control-rate DSP is planned.
This would entail Scheme executing in the timed audio rendering thread, but
at a subdivision of the audio sample rate such as every 32 or 64 samples, 
thus generating streams similar to the way digital audio is rendered, but
at a lower resolution suitable for use as signal modulators.

The author also plans to create portability layers in Scheme to allow one to run 
the same Scheme code on S4M, S4Pd, or Common Music (which also uses s7),
and potentially in C++ projects through the JUCE audio development framework.
There have also been successful ports of s7, Pure Data, and Csound to Web Assembly, 
thus it seems likely that work from the project will be useable for web audio as well.
It is hoped that through these additional run-times the project can become 
a valuable addition to the larger computer music community, enabling
use in contexts where the commercial Max platform is not practical
or financially accessible.

@@

https://github.com/iainctduncan/scheme-for-max
%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
Scheme For Max would not exist without the work of many people. 
Bill Schottstaedt created Common Lisp Music and the s7 Scheme implementation,
and has been invaluable as a source of advice and technical assistance. Heinrich Taube
created Common Music and has been an inspiration through his seminal book
on algorithmic composition with Lisp, ``Notes from the Metalevel".
Luigi Castelli and Alex Harker helped with Max external programming questions
and build assistance. 
S4M of course depends on Max and thus owes a huge debt to 
Miller Puckette and David Zicarelli.
George Tzanetakis, my supervisor, assisted with this paper.
And finally, Andrew Schloss, my other supervior, has
been a source of encouragement and support through the years of 
exploration that led to the Scheme For Max project.
